\documentstyle[12pt,A4]{article}
\begin{document}
\thispagestyle{blank}
{\LARGE Errata/Omissions in the {\em PEW} User Manual}

\mbox{}

{\large p.5 Release Contents and Installation}

\mbox{}

Two files have been added: {\em PEWDEMO.DAT} containing a
demonstration script for the {\em PEW}, and {\em DEMO.BAT}, a batch
file which runs the demonstration.

\mbox{}

{\large p.7 The Main Menu}

\mbox{}

From 4th last line ("If one of...") to last line ("...to select a
menu item.") is much clearer if read as "If one of the two submenus ({\em File} or
{\em Options}) is invoked, and subsequently exited by pressing {\em
ESC}, you will be placed on the main menu but the submenu will close.
You can then move around the main menu with the left and right
cursor keys, and press {\em ENTER} to select a main menu item."

\mbox{}

{\large p. 25 }

\mbox{}

Last paragraph, 3rd line: "This example..." to 4th last line
"...network body type." is incorrect. This sentence should read:
"This example shows three breakpoints: the first is set on the 
{\em ack\_wait $\rightarrow$ ack\_wait} transition of the {\em
alternating\_bit[1]} process being enabled;
the second is a breakpoint on an {\em Output}
statement to the {\em n} interaction point of all processes having
the {\em alternating\_bit} body type, and the third is a line
breakpoint set on line 165." Note that suspended
breakpoints are displayed with parentheses about their numbers in the
breakpoint browser.

\mbox{}

{\large p28. Some more figures}

\begin{itemize}
\item Program size: 145kB (+7kB errors + 30kB help)
\item {\em PEW} help screens: 98
\item Language help: 34 screens indexed on 46 keys
\item Maximum number of {\em failed} When clauses per process: 32
\end{itemize}

\mbox{}

{\large p36. Stack overflow}

\mbox{}

As E-Code is stack-based all variables are allocated on the stack. 
This should be borne in mind when declaring variables, as stack space
is limited.

\mbox{}

{\large p36. Too many failed WHEN clauses in process}

\mbox{}

The maximum number allowed of failed When clauses within a single
process is 32.
\end{document}

